---
title: "DS_modelling"
output: html_document
date: "2026-05-20"
---

COMMENT: This needs to be updated with the correct condition coding - currently, it is not correct!

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
install.packages("pacman")
```


```{r}
pacman::p_load(tidyverse,
               lme4,
               emmeans,
               DHARMa,
               lmerTest,
               RColorBrewer, 
               glmmTMB)
```

```{r}
# data
data <- read_csv("../curvefit/clustered_data_w_condition.csv")
```

```{r}
head(data)
```

```{r}
# pre-processing
# adding new variable with n clusters;
# we also need to add a condition variable; this can be extracted with simple regex from the file names
data <- data %>% 
  group_by(Filename) %>% 
  mutate(n_clusters = max(`Cluster ID`),
         n_words = n())

# reformatting condition + participant variable
data$Condition <- as.factor(data$Condition)
data$ID <- as.factor(data$ID)
```

Simple corr test to check if n_words and n_clusters are correlated:

```{r}
cor.test(data$n_clusters, data$n_words)
```
```{r}
plot(data$n_clusters, data$n_words)
```

They are! This largely makes m2 redundant.

```{r}
plot(data$n_clusters)
```
Testing the curve parameters

```{r}
cor.test(data$n_words, data$curve_c)
```

# Fitting VF models

```{r}
# transforming data to a single row for each unique participant+file combination 
data_transformed <- data %>%
  distinct(ID, Condition, n_clusters, n_words, curve_c, curve_m)

# fitting models using a poisson distribution for n_clusters
m1 <- glmer(formula = n_clusters ~ Condition + (1|ID), 
      data = data_transformed, family = poisson(link = "log"))

m2 <- glmer(formula = n_words ~ Condition + (1|ID), 
      data = data_transformed, family = poisson(link = "log"))

# fitting models to curve_c / curve_m
m3 <- lmer(formula = curve_c ~ Condition + (1|ID), 
      data = data_transformed)

m4 <- lmer(formula = curve_m ~ Condition + (1|ID), 
      data = data_transformed)
```

```{r}
# checking n rows in transformed data
nrow(data_transformed)
```


```{r}
summary(m1) # notice that coefficients are returned on the log scale; we need exp() to get them back to our scale
#summary(m2)
```
Confidence intervals:
```{r}
m1_conf <- confint(m1)
```
```{r}
round(m1_conf, 2)
```


Testing for overdispersion:

```{r}
simulationOutput <- simulateResiduals(m1)
testDispersion(simulationOutput)
```
Data is underdispersed, not overdispersed. We will not change the model based on this.


```{r}
emm_df <- as.data.frame(emmeans(m1, ~ Condition, type = "response"))
data_avg <- distinct(data, ID, Condition, n_clusters)

plot(jitter(as.numeric(data_avg$Condition), 0.3), data_avg$n_clusters,
     xaxt = "n", xlab = "Condition", ylab = "Number of Clusters",
     pch = 16, col = rgb(0, 0, 0, 0.3))

axis(1, at = 1:3, labels = c("1", "2", "3"))

lines(as.numeric(factor(emm_df$Condition)), emm_df$rate)

points(as.numeric(factor(emm_df$Condition)), emm_df$rate, pch = 16)

arrows(as.numeric(factor(emm_df$Condition)), emm_df$asymp.LCL,
       as.numeric(factor(emm_df$Condition)), emm_df$asymp.UCL,
       angle = 90, code = 3, length = 0.05)
```

```{r}
emm_df <- as.data.frame(emmeans(m1, ~ Condition, type = "response"))
data_avg <- data %>%
  distinct(ID, Condition, n_clusters) %>%
  group_by(ID, Condition) %>%
  summarise(n_clusters = mean(n_clusters), .groups = "drop")

id_list <- unique(data_avg$ID)
colors <- c("#E63946", "#2A9D8F", "#E9C46A", "#457B9D", "#F4A261",
            "#6A0572", "#2D6A4F", "#F77F00", "#4361EE", "#B5179E", "#06D6A0")
id_colors <- setNames(colors[1:length(id_list)], id_list)

plot(as.numeric(data_avg$Condition), data_avg$n_clusters,
     xaxt = "n", xlab = "Condition", ylab = "Number of Clusters",
     type = "n",
     main = "Number of Clusters By DBS Condition: Estimates and Data")
axis(1, at = 1:3, labels = c("DBS Off", "DBS On - Dorsal", "DBS On - Ventral"))

for (id in id_list) {
  d <- data_avg[data_avg$ID == id, ]
  d <- d[order(d$Condition), ]
  lines(as.numeric(d$Condition), d$n_clusters,
        type = "o", pch = 16,
        col = adjustcolor(id_colors[as.character(id)], alpha.f = 0.5))
}

lines(as.numeric(factor(emm_df$Condition)), emm_df$rate, lwd = 2)
points(as.numeric(factor(emm_df$Condition)), emm_df$rate, pch = 16)
arrows(as.numeric(factor(emm_df$Condition)), emm_df$asymp.LCL,
       as.numeric(factor(emm_df$Condition)), emm_df$asymp.UCL,
       angle = 90, code = 3, length = 0.05)
```

```{r}
# this can be used to report the results
emm_df 
```

# Checking model 2, 3 and 4

```{r}
summary(m2)
```


```{r}
summary(m3)
```
```{r}
plot(y = data_transformed$curve_c, x = as.numeric(data_transformed$Condition))
```
Ok, so there is a huge outlier here as well. Let's identify it:

```{r}
outlier <- data_transformed %>% filter(curve_c > 150) %>% pull(Filename)
print(outlier)
```

After looking at the curve fit plot for this file, there are no obvious problems with the fit. So no grounds for excluding it.


```{r}
summary(m4)
```
```{r}
plot(y = data_transformed$curve_m, x = data_transformed$Condition)
```
Ok, so this is driven by an extreme outlier; lets look at it.

```{r}
outlier <- data_transformed %>% filter(curve_m > 100) %>% pull(Filename)
print(outlier)
```

After inspecting the curve fit plot, the curve is clearly misspecified here; the curve is fit to a flat horizontal line, and all utterances are identified as belonging to one cluster. This can be discarded.

```{r}
data_transformed_no_outlier <- data_transformed %>% filter(!Filename==outlier)
```

Refitting model:

```{r}
m4.1 <- lmer(formula = curve_m ~ Condition + (1|ID), 
      data = data_transformed_no_outlier)
summary(m4.1)
```
```{r}
plot(y = data_transformed_no_outlier$curve_m, x = as.numeric(data_transformed_no_outlier$Condition))
```
But even without this outlier, there is still no obvious effect. 

# Assorted plots and statistics

```{r}
# number of clusters
table(data_transformed$n_clusters)
summary(data_transformed$n_clusters)
```

# Fitting linguistic models

```{r}
data_ling <- read_csv("../curvefit/clustered_data_w_linguistics.csv")
```

```{r}
# first, some transformations
data_ling <- data_ling %>%
  distinct(Filename, `Cluster ID`, .keep_all = TRUE)
```


```{r}
plot(y = data_ling$avg_sim, x = as.numeric(data_ling$Condition),
     main = "Avg. Semantic Similarity by DBS Condition",
     ylab = "Semantic Similarty",
     xlab = "Condition")
plot(y = data_ling$avg_phon_sim, x = as.numeric(data_ling$Condition),
     main = "Avg. Phonemic Similarity by DBS Condition",
     ylab = "Phonemic Similarty",
      xlab = "Condition")
```
```{r}
blue <- adjustcolor("steelblue", alpha.f = 0.2)
red  <- adjustcolor("tomato", alpha.f = 0.2)

plot(y = data_ling$avg_sim,
     x = as.numeric(data_ling$Condition) - 0.1,
     main = "Semantic & Phonemic Similarity by DBS Condition",
     ylab = "Similarity",
     xlab = "Condition",
     xaxt = "n",
     col = blue,
     pch = 16,
     ylim = c(0, 1),
     xlim = c(0.5, nlevels(data_ling$Condition) + 0.5))

points(y = data_ling$avg_phon_sim,
       x = as.numeric(data_ling$Condition) + 0.1,
       col = red,
       pch = 16)

axis(1, at = 1:nlevels(data_ling$Condition), labels = levels(data_ling$Condition))

legend("bottomleft",
       legend = c("Semantic", "Phonemic"),
       col = c(blue, red),
       pch = 16,
       cex = 0.8)
```

```{r}
colnames(data_ling)
```
```{r}
# transforming phon sim data to not hit the bounds
data_ling_trans <- data_ling %>% 
  mutate(avg_phon_sim = case_when(
    avg_phon_sim == 0 ~ 0.001,        
    avg_phon_sim == 1 ~ 0.999,         
    TRUE ~ avg_phon_sim            
  ))
```

```{r}
# checking whats going on with extreme scores actually!
data_ling %>% filter(avg_phon_sim > 0.9)
```

```{r}
# checking the number of rows prior to fitting
nrow(data_ling_trans)
```


```{r}
# fitting models using a poisson distribution for n_clusters
# semantic similarity model
m5 <- glmmTMB(formula = avg_sim ~ Condition + (1|ID), 
      data = data_ling, family = beta_family(link = "logit"))

# phonemic similarity model
m6 <- glmmTMB(formula = avg_phon_sim ~ Condition + (1|ID), 
      data = data_ling_trans, family = beta_family(link = "logit"))
```

```{r}
summary(m5)
```
Computing confidence intervals:

```{r}
round(confint(m5, level = 0.95), 2)
```
```{r}
summary(m6)
```
```{r}
round(confint(m6, level = 0.95), 2)
```


```{r}
# we can figure out the change in similarity like this:
plogis(0.55390)
plogis(0.55390 + 0.19580)
plogis(0.55390 + 0.20602)

```










